Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-11 paper/paper.tex
update

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{element type}     & \textbf{property to compute} \\
   {$\mathbbm{R}$}     & {the number of independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {MIS size} \\
   {Polynomial truncated to highest order (\Eq{eq:countingtropical})}     & {MIS size and the number of MISs} \\
   {Polynomial truncated to 2nd highest order} (\Eq{eq:max2poly})     & {the number of MISs and independent sets of size $\alpha(G)-1$} \\
   {Set} (\Eq{eq:set})     & {enumeration of independent sets} \\
   {Bit string} (\Eq{eq:singleconfig})     & {one independent set} \\
   {Polynomial truncated to highest order combined with Bit string}     & {MIS size and one MIS} \\
   {Polynomial truncated to highest order combined with Set} (\Eq{eq:countingtropicalset})    & {MIS size and all MISs} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

This set algebra can serve as the coefficients in \Eq{eq:polynomial} to enumerate independent sets of all different sizes, \Eq{eq:countingtropical} to enumerate all MISs,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial forms a commutative semiring, the polynomial itself is still a commutative semiring.
For example, to enumerate only the MISs, with the tropical algebra, we can define $s_{k}\infty^k$,
where the coefficients follow the algebra in \Eq{eq:set} and the orders (exponents) follow the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm P1+SN}(\{\boldsymbol{e}_{i}\infty^1\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^1
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm P1+SN} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

@@

\subsection{Bounding the MIS enumeration space}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. In these graphs, a circle is a tensor, an arrow specifies execution direction of a function and $\circ$ is the Hadamard (element-wise) multiplication. $\overline A$ means the boolean mask of $A$. (a) is the forward pass with algebra \Eq{eq:tropical} for computing $\alpha(G)$.
     (b) is the backward pass for computing boolean gradients as masks.
     (c) is the masked forward pass with algebra \Eq{eq:countingtropicalset} for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

\subsection{Extremely large MIS enumeration space}
In many times, the number of configurations is too large to fit into any type of storage.
Then instead of using the set algebra, we can use a binary sum-product tree as a compact representation of the results.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
$\{\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM}, \texttt{PROD}\}$, $data$ is a bit string as the content is a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.

\begin{equation}
\eqname{EXPR}
\begin{split}
    s \oplus t &= (\texttt{SUM}, /, s, t)\\
    s \odot t &= (\texttt{PROD}, /, s, t)\\
    \mymathbb{0} &= (\texttt{ZERO}, /, /, /)\\
    \mymathbb{1} &= (\texttt{LEAF}, 0^{\otimes |V|}, /, /).
\end{split}\label{eq:expr}
\end{equation}

@@

\begin{equation}
\eqname{Tk}
\begin{split}
    s \oplus t &= \texttt{largest}(s \cup t, k)\\
    s \odot t &= \texttt{largest}(\{a+b| a \in s, b\in t\}, k)\\
    \mymathbb{0} &= -\infty^{\otimes k}\\
    \mymathbb{1} &= -\infty^{\otimes k-1} \otimes 0
\end{split}\label{eq:ext-tropical}
\end{equation}
where $\texttt{largest}(s, k)$ means truncating the set by only keeping $k$ largest values in the set $s$.
Similarly, we can attach a bit string sampler algebra \Eq{eq:singleconfig} to it for finding the configuration for each size.
Note here, the $\odot$ of configuration sampler is not used so that the resulting configurations are complete.

@@

\begin{figure}[th]
   \centering
\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[draw,fill=lime!80,fill opacity=1, text opacity=1.0,ellipse,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    \def\dx{-3};
    \node[draw,fill=teal!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[draw,fill=brown!75,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.2cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-0.7cm}Highest order};
    \node[draw,fill=brown!40,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=0.7cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.7cm}2nd highest order};
    \node[draw,fill=brown,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above of=2] at (T) (textT) {Tropical};
    \draw[black,-latex] (textT) -- (T);

@@

>>>>

\begin{table}[t!]\centering
\begin{minipage}{\columnwidth}
\ra{1.3}
        \begin{tabularx}{\textwidth}{sb}\toprule
            \hline
   \textbf{Element type}     & \textbf{Property to compute} \\
   {$\mathbbm{R}$}     & {Counting of all independent sets} \\
   {Polynomial} (\Eq{eq:polynomial})     & {Independence polynomial} \\
   {Tropical (\Eq{eq:tropical})}    & {Maximum independent set size} \\
   {Extended Tropical of order $k$ (\Eq{eq:ext-tropical})}    & {Largest $k$ independent set sizes} \\
   {Polynomial truncated to $k$-th order (\Eq{eq:countingtropical} and \Eq{eq:max2poly})}     & {$k$ largest independent sizes and their counting} \\
   {Set} (\Eq{eq:set})     & {Enumeration of independent sets} \\
   {Sum Product Tree} (\Eq{eq:expr})     & {Sampling of independent sets} \\
   {Polynomial truncated to highest order combined with Bit string} (\Eq{eq:singleconfig})     & {Maximum independent set size and one of such configurations} \\
   {Polynomial truncated to $k$-th order combined with Set} (\Eq{eq:countingtropicalset})    & {$k$ largest independent set sizes and their enumeration} \\
            \bottomrule
        \end{tabularx}
    \caption{Tensor element types and the independent set properties that can be computed using them.}\label{tbl:generictypes}
\end{minipage}
\end{table}

@@

This set algebra can serve as the coefficients in \Eq{eq:polynomial} to enumerate independent sets of all different sizes, \Eq{eq:countingtropical} to enumerate all MISs,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial forms a commutative semiring, the polynomial itself is still a commutative semiring.
For example, to enumerate only the MISs, with the tropical algebra, we can define $s_{k}\infty^k$,
where the coefficients follow the algebra in \Eq{eq:set} and the orders (exponents) follow the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm P1+SN}(\{\boldsymbol{e}_{i}\infty^1\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^1
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm P1+SN} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.
Direct implementing this algebra for configuration MISs might incur significant space overheads for keeping too many intermediate states not contributing to the final maximum independent sets,
one can avoid this issue by using the bounding technique in \App{sec:bounding}.

@@

\subsection{Sampling extremely large configuration space}
In many times, the number of configurations is too large to fit into any type of storage.
Then instead of using the set algebra, we can use a binary sum-product tree as a compact representation of the results.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
$\{\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM}, \texttt{PROD}\}$, $data$ is a bit string as the content is a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.

\begin{equation}
\eqname{SUMPROD}
\begin{split}
    s \oplus t &= (\texttt{SUM}, /, s, t)\\
    s \odot t &= (\texttt{PROD}, /, s, t)\\
    \mymathbb{0} &= (\texttt{ZERO}, /, /, /)\\
    \mymathbb{1} &= (\texttt{LEAF}, 0^{\otimes |V|}, /, /).
\end{split}\label{eq:expr}
\end{equation}
The vertex tensor and edge tensor become
\begin{equation}
    W^{\rm SUMPROD}((\texttt{LEAF},\boldsymbol{e}_i,/,/)) = \left(\begin{matrix}
        \mymathbb{1} \\
        (\texttt{LEAF},\boldsymbol{e}_i,/,/)
    \end{matrix}\right),   
    \qquad
        B^{\rm SUMPROD} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

\begin{equation}
\eqname{Tk}
\begin{split}
    s \oplus t &= \texttt{largest}(s \cup t, k)\\
    s \odot t &= \texttt{largest}(\{a+b| a \in s, b\in t\}, k)\\
    \mymathbb{0} &= -\infty^{\otimes k}\\
    \mymathbb{1} &= -\infty^{\otimes k-1} \otimes 0
\end{split}\label{eq:ext-tropical}
\end{equation}
where $\texttt{largest}(s, k)$ means truncating the set by only keeping $k$ largest values in the set $s$.
The vertex tensor and edge tensor become
\begin{equation}
    W^{\rm Tk}(-\infty^{\otimes k-1} \otimes 1) = \left(\begin{matrix}
        \mymathbb{1} \\
        -\infty^{\otimes k-1} \otimes 1
    \end{matrix}\right),   
    \qquad
        B^{\rm Tk} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

Similar to the Tropical numbers, we can attach a bit string sampler algebra \Eq{eq:singleconfig} to it for finding the configurations of each size.
Note here, the $\odot$ of configuration sampler is not used so that the resulting configurations are complete.

@@

\begin{figure}[th]
   \centering
\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[draw,fill=lime!80,fill opacity=1, text opacity=1.0,ellipse,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    \def\dx{-3};
    \node[draw,fill=teal!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[draw,fill=brown!75,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.0cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-1.0cm}Highest order};
    \node[draw,fill=brown!40,fill opacity=1, text opacity=1.0,ellipse,minimum width=3.5cm, minimum height=1.0cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.8cm}2nd highest order};
    \node[draw,fill=brown,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node[draw,fill=brown!50,fill opacity=1, text opacity=1.0,ellipse,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, -0.1) (T2) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above = 1cm] at (T) (textT) {Tropical};
    \node[below = 2cm, left=4cm] (textT2) {Extended Tropical};
    \draw[black,-latex] (textT) -- (T);
    \draw[black,-latex] (textT2) -- (T2);

@@

\section{Bounding the MIS enumeration space}\label{sec:bounding}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. In these graphs, a circle is a tensor, an arrow specifies execution direction of a function and $\circ$ is the Hadamard (element-wise) multiplication. $\overline A$ means the boolean mask of $A$. (a) is the forward pass with algebra \Eq{eq:tropical} for computing $\alpha(G)$.
     (b) is the backward pass for computing boolean gradients as masks.
     (c) is the masked forward pass with algebra \Eq{eq:countingtropicalset} for enumerating configurations.}
     \label{fig:bounding}
\end{figure}
